---
title: "Model Fitting Basics"
author: "Christopher Conley, Pei Wang, Jie Peng"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Fitting Basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(message=F, warning=F)
```

<!--## Motivation -->

This vignette helps users with the network fitting function  [spacemap](https://topherconley.github.io/spacemap/reference/spacemap.html) and its output. 
<!--Familiarity with the mathematical expression of the spaceMap model will help the reader map the notation to R object variables (see the Methods section of the spacemap publication).  -->

## Background 

The spaceMap model is a conditional graphical model  for learning networks with two-types of nodes. Two distinct data (node) types are input to the model: One labeled as $X$'s (predictor variables) that can be continuous, discrete, or categorical in nature; another  labeled as $Y$'s that are continuous response variables. <!--When we fit the spaceMap model, we seek to learn the network comprised of node set $\textbf{V}= (X,Y)$ and edge set $\textbf{E}= \{ x_p - y_q : p \in \{1, \dots, P\}, q \in \{1, \dots, Q\} \} \cup \{ y_q - y_l : q \neq l \in \{1, \dots, Q\} \}$. -->

The output of the spaceMap model is a network containing  $x-y$ and $y-y$ edges, but not $x-x$ edges. 
<!--Motivated by integrative genomics applications, it has been applied to learn how genomic copy number alterations (i.e. $x$'s) regulate protein expression co-variation (i.e. $y$'s) in breast cancer. -->
It has been used to identify which genomic alterations ($x$) disrupt which protein ($y$) pathways, while simultaneously accounting for the  interactions among the proteins (https://topherconley.github.io/neta-bcpls/). The model may also be applied  to other high dimensional data types and to applications beyond integrative genomics. 

## Example 

We use  simulation data [sim1](https://topherconley.github.io/spacemap/reference/sim1.html) to illustrate how to fit the  spaceMap model. [sim1] contains $N=150$ i.i.d. samples of $P=14$ predictor variables ($X$) and $Q=171$ response variables ($Y$) generated under a Gaussian Graphical Model. <!--that is to say, $ (X,Y) \sim N_{P + Q}(0,\Theta^{-1})$, where the non-zero elements of the inverse covariance matrix $\Theta$ specify edges between $x-y$ and $y-y$ and (and $x-x$, although these are ignored in the conditional framework). Normality is not a required model assumption for either predictor or response, but is useful for simulations because edges are easily parameterized through $\Theta$.--> The corresponding network has 4 prominent $x$-hubs perturbing ~13 $y$ variables, and two prominent $Y$-hubs. More details of this simulation can be found in the [sim1](https://topherconley.github.io/spacemap/reference/sim1.html) documentation. 


We load the `sim1` R object, containing the $N\times Q$ response matrix $\bf Y$, the $N\times P$ response matrix $\bf X$, and  the true network.  

```{r}
library(spacemap)
data(sim1)
```

<!--The spaceMap model learns the structure of the network from data by estimating the edge set $\bf\hat E$.--> The following steps will discuss input parameters to learn the network and the output.  

## Fitting `spacemap`

### Tuning parameters

Tuning parameters $\lambda_1, \lambda_2$ (corresponding to the arguments `lam1` and `lam2`) control the overall sparsity of the $y-y$ and $x-y$ edges in the network, respectively. Tuning penalty $\lambda_3$ (corresponding to argument `lam3`) encourages selection of $x$ hubs-- $x$ that influences many $y$'s. 

The selection of tuning parameters through cross-validation is discussed in another [vignettes](https://topherconley.github.io/spacemap/articles/tuning.html). Here,  we set the tuning parameters based on previous cross validation results. 

```{r}
lam1 <- 72
lam2 <- 29 
lam3 <- 18
```

### Output

<!--Fit the spacemap model to the data matrices $\bf X$ and $\bf Y$ with the selected tuning penalties.--> All other input parameters are left as defaults, but can be further customized as needed (see `help(spacemap::spacemap)`).  

```{r}
net <- spacemap(Y = sim1$Y, X = sim1$X, lam1 = lam1, lam2 = lam2, lam3 = lam3)
```

The network topology is encoded in the adjacency matrices `yy` and `xy` output from the `adjacency` auxiliary function. 

```{r}
adjnet <- adjacency(net = net)
str(adjnet)
```

where 

+ `adjnet$yy[q,l]` indicates an edge $y_q - y_l$ when equal to 1 and 0 otherwise. 
+ `adjnet$xy[p,q]` indicates an edge $x_p - y_q$ when equal to 1 and 0 otherwise. 

The output of the `spacemap` function is a list of estimated parameters and model diagnostics.
<!--The estimated parameters are combined to infer the network. The model diagnostics help the user to decide whether the estimated parameters are valid.-->

```{r}
str(net)
```

##### Estimated parameters:

+ `ParCor`: The estimated $y-y$ partial correlation matrix (given $x$). 
+ `sig.fit`: The estimated diagonal of the $y-y$ submatrix of the inverse joint covariance matrix.
+ `Gamma`: The estimated coefficients for $x-y$ edges. 

##### Model Diagnostics: 

+ `rss`: numerica,  the residual sum of squares. 
+ `convergence`: logical,  indicating whether the model converged subject to the specified tolerance and number of maximum iterations; 
+ `deltaMax`: numeric, the maximum change in parameter values in the last iteration. 


## Visualization

There are many software for network visualization, e.g.,  *Cytoscape* (http://cytoscape.org/index.html). <!--However, such a tool may involve some manual curation and the sophisticated options makes for a steeper learning curve.--> The following section illustrates how to visualize the network output using the `igraph' and  `networkD3` R packages. 

First, export the the spacemap output to an igraph object which  has a sparse representation.

```{r}
library(igraph)
ig <- spacemap::adj2igraph(yy = adjnet$yy, xy = adjnet$xy)
```

<!--with `ig` occupying `r round(object.size(adjnet) / object.size(ig),0)` times smaller memory than `adjnet`.-->  

We then identify network modules with igraph's edge-betweenness algorithm.

```{r}
ceb <- igraph::cluster_edge_betweenness(graph = ig)
members <- igraph::membership(ceb)
```
 
<!--Color the nodes according to their communities. Since the community detection method we used was hierarchical, we can further group higher-order structure of the communities with shaded polygons. -->
We then plot the network where edges between modules are colored red and edges  within modules are colored in black, and nodes with higher degree are scaled to be larger. 

```{r}
dg <- degree(ig)*1.5
plot(ceb, ig, vertex.label= "", vertex.size = dg,
edge.width = rep(2, ecount(ig)))
```

For interactive visualization, we could use the `networkD3` package. The following snippet will visualize the network where $X$ and $Y$ hub nodes are larger in size and different colors denote the 14 different modules. Hover over a node with mouse if you wish to see the node ID and its neighbors. 

 
```{r}
library(networkD3)
#Convert the object suitable for `networkD3`
ig_d3 <- igraph_to_networkD3(ig, group = members)
#make the node size bigger according to degree
ig_d3$nodes$degree <- igraph::degree(ig)^2
forceNetwork(Links = ig_d3$links, Nodes = ig_d3$nodes, 
             linkDistance = 10,
             Source = 'source', Target = 'target',
             fontSize = 12,
             NodeID = 'name', Group = 'group', Nodesize = "degree", 
             opacity = 0.90, 
             bounded = T, 
             charge = -15)
```


## Further Reading

With this basic understanding of the `spacemap` model fitting, we recommend looking at the [next vignette](https://topherconley.github.io/spacemap/articles/tuning.html) which illustrates how to select tuning parameters through an iterative grid-search with cross-validation criterion. 

